Assessing primitives performance\\ on multi-stage execution

### 2018-05-18 paper.tex
make things consistent

\def\legBase{Baseline\xspace}
\def\legSmartSyntaxPlug{Syntax\xspace}
\def\legSlangPlug{Plugin\xspace}
\def\legSlang{Slang\xspace}
\def\legRTL{Slang+RTL\xspace}
\def\plotHeight{7cm}

@@

\paragraph{GCC-Linux.}The GCC-Linux evaluation was performed on an Asus Zenbook with Ubuntu 16.04 LTS, a 2.2 Ghz processor Intel Core 5, 8 Gb 1600MHz DDR3 of RAM. The Linux VM was compiled using GCC version 5.4.0.

\paragraph{LLVM-Mac.}The LLVM-Mac evaluation was performed on a MacBook pro with Mac OS 10.11.6, a 2.9 Ghz processor Intel Core 5, 8 Gb 1867MHz DDR3 of RAM. The Mac VM was compiled using Apple LLVM version 8.0.0 (clang-800.0.38).

@@

   \subcaption*{\textbf{GCC-Linux}\vspace{0.25cm}}

@@

   \end{subfigure}
   \subcaption*{\textbf{LLVM-Mac}\vspace{-0.2cm}} %change vspace based on fig position
    \end{subfigure}
   \caption{String comparison benchmarks for strings of different sizes on different architectures.}
   \label{fig:bench}
\end{figure*}
\begin{table*}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|}
  \hline
    & & 0 & 1 & 5 & 100 & 1000 \\
\hline
  \multirow{2}{*}{\legSlangPlug} &GCC-Linux& 0.4222 $\pm$ 0.0008 & 0.7674 $\pm$ 0.0011 & 1.8906 $\pm$ 0.0119 & 13.1218 $\pm$ 0.0171  & 25.3595 $\pm$ 0.1742 \\
  						&LLVM-Mac& 0.3856 $\pm$ 0.0004 & 0.6685 $\pm$ 0.0010 & 1.6387 $\pm$ 0.0101 & 11.6376 $\pm$ 0.0576  & 23.9579 $\pm$ 0.1390 \\
   \hline
  \multirow{2}{*}{\legSlang} &GCC-Linux& 1.0288 $\pm$ 0.0046 & 1.8431 $\pm$ 0.0116 & 4.0412 $\pm$ 0.0611 & 17.7084 $\pm$ 0.1059 & 25.4258 $\pm$ 0.1621\\
 					 &LLVM-Mac& 0.2088 $\pm$ 0.0002 & 0.3649 $\pm$ 0.0064 & 0.8953 $\pm$ 0.0056 & 6.1978 $\pm$ 0.0180  & 10.6698 $\pm$ 0.0066 \\
   \hline
  \multirow{2}{*}{\legRTL}  &GCC-Linux& 2.4290 $\pm$ 0.0029 & 3.5170 $\pm$ 0.0197  & 7.1216 $\pm$ 0.0105 & 18.5537 $\pm$ 0.0238 & 26.7626 $\pm$ 0.1861\\
 				 	&LLVM-Mac & 2.5718 $\pm$ 0.0035 & 3.5919 $\pm$ 0.0076 & 6.9569 $\pm$ 0.0416 & 17.2747 $\pm$ 0.1071  & 25.3991 $\pm$ 0.1811 \\
  \hline
\end{tabular} 
\vspace{0.2cm}
\caption{Benchmark results with standard errors. Average relative speed-up regarding baseline.}
\label{tab:detail_bench}
\end{table*}

\begin{figure*}[b!]
		\centering
		\includegraphics[width=1\linewidth]{figures/asm_cmp.pdf}
		\caption{GCC and LLVM: different optimizations on the comparison loop.}
		\label{fig:Asm}
\end{figure*}

@@

\begin{code}
#(0 1 5 100 1000) collect: [ :size |
	| iterations time overhead |
	collection := ByteString new: size.
	collection2 := ByteString new: size.
	size < 100 
		ifTrue: [iterations := (100000000 // (size*10 max: 1) sqrt) floor.]
		ifFalse: [iterations := (10000000 // size sqrt) floor.].
	overhead := [ 1 to: iterations do: [ :i | ] ] timeToRun.
	time := [ 1 to: iterations do: [ :i |
		collection compareWithSlang: collection2 ] ] timeToRun.
	stream nextPutAll: ((time - overhead) asString)].
\end{code}\\

@@

The loop overhead is measured and removed from the measurements. The two compared strings are equal as this situation corresponds to the worst scenario for most implementations. %There is one less branch in the two plugin implementations in this case.

\subsection{Results} 
We performed the benchmarks upon five different string sizes: 0, 1, 5, 100 and 1000 characters. In the case of small strings, most of the time is spent activating the primitive and in the code before the main comparison loop, the primitive is performed with a little overall number of native instructions. In the case of large strings, the time spent executing the primitive is dominated by the main comparison loop. Figure~\ref{fig:bench} shows the results in terms of relative speed-up towards the baseline implementation, respectively for GCC-Linux and LLVM-Mac.

We can draw the following patterns from these results:
\begin{itemize}
\item The \legSlangPlug implementation is slower than the baseline version for small strings (0,58 times slower for empty strings, 0,23 times slower for 1-character strings). This is due to the overhead of the calls to the main VM to access the object headers and the stack. On large strings, the performance is comparable to 
\item The \legRTL implementation is more efficient than the baseline in every case. Nevertheless, the performance gap between this version and the \legSlangPlug and \legSlang ones tends to decrease as the string size increases, especially for GCC. For 1000-characters strings, the \legRTL version goes 0,25 times faster than the \legSlangPlug and the \legSlang versions.
\item The compiler used (either GCC or LLVM) does lead to significant differences in terms of performance for the Slang implementation. 
\end{itemize}

@@

\paragraph{Stack switch overhead.}
As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks. This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is between 57,8 (gcc) to 61,4\% (llvm) slower than the baseline's for empty strings.
As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
Regarding this criteria, \legRTL version is always the fastest as it is not impacted by the switch overhead.

\paragraph{Optimization potential on large strings comparison.}
The Plugin come with an overhead due the use of an interface. Besides, cross-file inlining is not performed for plugin, like said in Section~\ref{sec:implem_cmp}. These are some of the factors explaining the performance gap between these versions and the \legSlang and \legRTL ones.
When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure~\ref{fig:Asm}). However, the switch overhead is still noticeable in the Slang implementation when strings of 100 characters are compared: Slang is 17.7 times faster than Baseline while Slang+JIT is 18,6 times faster than Baseline. 
This last observation only applies to the GCC results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph.

\paragraph{Control over generated code.} One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performance between the Slang and Slang+JIT implementations is really close for long strings (100 characters and longer), with respectively 25,42 and 26,76 relative speed-up factors for a 1000-characters string. Yet performance greatly differs when LLVM is used, with respectively 10,67 and 25,4 relative speed-up factors.

@@

In this section, we compared the performance of the different implementations against a baseline (\emph{i.e.,}{pure Smalltalk code}) implementation. The results showed the Slang+JIT implementation was always the fastest. Otherwise, most of the performance gaps between the primitive versions could be explained by 3 criteria: stack swithing, optimization for large strings, and the control over generated code.

>>>>

\def\legBase{Baseline\xspace}
\def\legSmartSyntaxPlug{Syntax\xspace}
\def\legSlangPlug{Plugin\xspace}
\def\legSlang{Slang\xspace}
\def\legRTL{Slang+RTL\xspace}
\def\LLVMMac{LLVM-Mac\xspace}
\def\GCCLinux{GCC-Linux\xspace}
\def\plotHeight{7cm}

@@

\paragraph{\GCCLinux.}The \GCCLinux evaluation was performed on an Asus Zenbook with Ubuntu 16.04 LTS, a 2.2 Ghz processor Intel Core 5, 8 Gb 1600MHz DDR3 of RAM. The Linux VM was compiled using GCC version 5.4.0.

\paragraph{\LLVMMac.}The \LLVMMac evaluation was performed on a MacBook pro with Mac OS 10.11.6, a 2.9 Ghz processor Intel Core 5, 8 Gb 1867MHz DDR3 of RAM. The Mac VM was compiled using Apple LLVM version 8.0.0 (clang-800.0.38).

@@

   \subcaption*{\textbf{\GCCLinux}\vspace{0.25cm}}

@@

   \end{subfigure}
   \subcaption*{\textbf{\LLVMMac}\vspace{-0.2cm}} %change vspace based on fig position
    \end{subfigure}
   \caption{String comparison benchmarks for strings of different sizes on different architectures.}
   \label{fig:bench}
\end{figure*}
\begin{table*}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|}
  \hline
    & & 0 & 1 & 5 & 100 & 1000 \\
\hline
  \multirow{2}{*}{\legSlangPlug} &\GCCLinux& 0.422 $\pm$ 0.0008 & 0.767 $\pm$ 0.0011 & 1.891 $\pm$ 0.0119 & 13.12 $\pm$ 0.0171  & 25.36 $\pm$ 0.1742 \\
  						&\LLVMMac& 0.386 $\pm$ 0.0004 & 0.669 $\pm$ 0.0010 & 1.639 $\pm$ 0.0101 & 11.64 $\pm$ 0.0576  & 23.96 $\pm$ 0.1390 \\
   \hline
  \multirow{2}{*}{\legSlang} &\GCCLinux& 1.029 $\pm$ 0.0046 & 1.843 $\pm$ 0.0116 & 4.041 $\pm$ 0.0611 & 17.71 $\pm$ 0.1059 & 25.43 $\pm$ 0.1621\\
 					 &\LLVMMac& 0.209 $\pm$ 0.0002 & 0.365 $\pm$ 0.0064 & 0.895 $\pm$ 0.0056 & 6.198 $\pm$ 0.0180  & 10.67 $\pm$ 0.0066 \\
   \hline
  \multirow{2}{*}{\legRTL}  &\GCCLinux& 2.429 $\pm$ 0.0029 & 3.517 $\pm$ 0.0197  & 7.122 $\pm$ 0.0105 & 18.55 $\pm$ 0.0238 & 26.76 $\pm$ 0.1861\\
 				 	&\LLVMMac & 2.572 $\pm$ 0.0035 & 3.592 $\pm$ 0.0076 & 6.957 $\pm$ 0.0416 & 17.28 $\pm$ 0.1071  & 25.40 $\pm$ 0.1811 \\
  \hline
\end{tabular} 
\vspace{0.2cm}
\caption{Benchmark results with standard errors. Average relative speed-up regarding baseline.}
\label{tab:detail_bench}
\end{table*}

\begin{figure*}[b!]
		\centering
		\includegraphics[width=1\linewidth]{figures/asm_cmp.pdf}
		\caption{\GCCLinux and \LLVMMac: different optimizations on the comparison loop.}
		\label{fig:Asm}
\end{figure*}

@@

\begin{code}
#(0 1 5 100 1000) collect: [ :size |
	| iterations time overhead |
	collection := ByteString new: size.
	collection2 := ByteString new: size.
	size < 100 
		ifTrue: [iterations := (100000000 // (size*10 max: 1) sqrt) floor.]
		ifFalse: [iterations := (10000000 // size sqrt) floor.].
	overhead := [ 1 to: iterations do: [ :i | ] ] timeToRun.
	time := [ 1 to: iterations do: [ :i |
		collection primitiveStringCompare: collection2 ] ] timeToRun.
	stream nextPutAll: ((time - overhead) asString)].
\end{code}\\

@@

The loop overhead is measured and removed from the measurements (We made sure the JIT did not remove the dead loop to measure it). The two compared strings are equal as this situation corresponds to the worst scenario for most implementations. %There is one less branch in the two plugin implementations in this case.

\subsection{Results} 
We performed the benchmarks upon five different string sizes: 0, 1, 5, 100 and 1000 characters. In the case of small strings, most of the time is spent activating the primitive and in the code before the main comparison loop, the primitive is performed with a little overall number of native instructions. In the case of large strings, the time spent executing the primitive is dominated by the main comparison loop. Figure~\ref{fig:bench} shows the results in terms of relative speed-up towards the baseline implementation, respectively for \GCCLinux and \LLVMMac.

\cba{Is next paragraph good ?}
We can draw the following patterns from these results:
\begin{itemize}
\item The \legSlangPlug implementation is slower than the baseline version for small strings (0,58 times slower for empty strings, 0,23 times slower for 1-character strings). This is due to the overhead of the calls to the main VM to access the object headers and the stack. On large strings, \legSlangPlug is one of the fastest implementation with 
\item The \legRTL implementation is more efficient than the baseline in every case. Nevertheless, the performance gap between this version and the \legSlangPlug and \legSlang ones tends to decrease as the string size increases, especially for \GCCLinux. For 1000-characters strings, the \legRTL version goes 0,25 times faster than the \legSlangPlug and the \legSlang versions.
\item The set-up used (either \GCCLinux or \LLVMMac) does lead to significant differences in terms of performance for the Slang implementation. 
\end{itemize}

@@

\paragraph{Stack switch overhead.}
As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks. This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is between 57,8 (\GCCLinux) to 61,4\% (\LLVMMac) slower than the baseline's for empty strings.
As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
Regarding this criteria, \legRTL version is always the fastest as it is not impacted by the switch overhead.

\paragraph{Optimization potential on large strings comparison.}
The Plugin come with an overhead due the use of an interface. Besides, cross-file inlining is not performed for plugin, like said in Section~\ref{sec:implem_cmp}. These are some of the factors explaining the performance gap between these versions and the \legSlang and \legRTL ones.
When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure~\ref{fig:Asm}). However, the switch overhead is still noticeable in the Slang implementation when strings of 100 characters are compared: Slang is 17.7 times faster than Baseline while Slang+JIT is 18,6 times faster than Baseline. 
This last observation only applies to the \GCCLinux results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph.

\paragraph{Control over generated code.} One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performance between the \legSlang and \legRTL implementations is really close for long strings (100 characters and longer), with respectively 25,42 and 26,76 relative speed-up factors for a 1000-characters string. Yet performance greatly differs when LLVM is used, with respectively 10,67 and 25,4 relative speed-up factors.

@@

In this section, we compared the performance of the different implementations against a baseline (\emph{i.e.,}{pure Smalltalk code}) implementation. The results showed the \legRTL implementation was always the fastest. Otherwise, most of the performance gaps between the primitive versions could be explained by 3 criteria: stack swithing, optimization for large strings, and the control over generated code.
